Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array: inclusion of pulsar terms

### 2023-11-15 manuscript/manuscript_2/paper.tex
attempt tp reproduce the earth/psr run 1251

The modulation in the pulsar TOAs due to the presence of a GW has two contributions: an ``Earth term" and a ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases. Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}. However, dropping the pulsar terms leads to biases in the inferred parameters and reduced detection probabilities \citep{Zhupulsarterms,Chen2022,KimpsonPTA}. \newline

\cite{KimpsonPTA} (K24 hereafter) introduced a novel, state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, that self-consistently tracks the intrinsic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations. The method described in K24 follows the example of standard  PTA analyses and also drops the pulsar term. In this paper we undertake two key tasks. Firstly we extend the methods presented in K24 to include the pulsar term. We show how to reparametrise the model in terms of an additional pulsars-specific phase term. We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}. Secondly we quantify how the inclusion of the pulsar term affects the accuracy with which the static GW parameters can be estimated, and the minimal detectable GW strain.

This paper is organised as follows. In Section \ref{sec:2} we briefly review the state-space formulation of PTA data analysis introduced in \cite{KimpsonPTA}.

Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

%In Section 3 we introduce how to include the pulsar term in the state-space model. In Section 4 we test the updated method on the synthetic data for a single representative example GW system, and compare the results to the original Earth-terms model. In Section 9 we extend the tests to cover an astrophysically relevant parameter space. Discussion on future extensions and conclusions are made in Sections 11 and 12 respectively. Throughout the paper we adopt the natural units, with 𝑐 = 𝐺 = ħ = 1, and metric signature (−, +, +, +).

%
%
%Intrinsic pulsar timing noise -- i.e.  random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Myers2021MNRAS.502.3113M,Meyers2021,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs), which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}. In modern Bayesian PTA searches, the power spectral density of the red intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the noise induced by a stochastic GW background (whose spectrum is also red). In addition to the red timing noise there are secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

@@

%The paper is organised as follows. In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW. In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence. In Section \ref{sec:testing} we describe how we create synthetic validation data with which to test the method. We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant domain of SMBHB source parameters. In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates. Conclusions are drawn in Section \ref{sec:discussion}.

@@

There are $N$ pulsars in the array. The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$. The intrinsic spin frequency of the $n$-th pulsar as measured in the local, freely-falling rest frame of the pulsar's centre of mass, $f_{\rm p}^{(n)}(t)$, evolves according to a stochastic differential equation. The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation which quantifies the influence of the GW. In Section \ref{sec:psr_frequency} we outline the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:psr_measured} we outline the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:ss_params} we summarise the parameters of the model. The model is formulated in terms of pulse frequency time series as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data. We defer this subtle generalization to future work.

\subsection{Spin evolution} \label{sec:psr_frequency}
The intrinsic pulsar spin frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ . 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes noise amplitude and the gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. \newline

@@

\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. We refer the reader to Appendix X for a review of Kalman filtering and nested sampling in the context of PTA searches for GWs. \newline

@@

\subsection{Reparametrisation of the pulsar term} \label{sec:pulsar_term2}
We define a new parameter 
\begin{equation}
	\chi^{(n)} = \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} 
\end{equation}
Such that Equation \eqref{eq:g_func_trig} becomes
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \chi^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig_chi}
\end{align}
and the static pulsar parameters to be inferred become
\begin{equation}
	\boldsymbol{\theta'}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\chi^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
\end{equation}
whilst $\boldsymbol{\theta}_{\rm gw}$  remain unchanged. We define $\boldsymbol{\theta'} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta'}_{\rm gw}$. \newline

@@

\section{Representative example: single SMBHB source and Comparison with Earth terms}\label{sec:rep_example1}

In this section we apply the analysis scheme of K23 and the new model parametrisation of Section \ref{sec:pulsar_term} to a PTA which is perturbed by a GW from an individual quasi-monochromatic SMBHB source. In Section \ref{sec:synthetic_data} we briefly review how synthetic data is generated in order to validate the analysis scheme. In Section \ref{sec:param_estimation} we apply the analysis scheme in conjunction with the model of Section \ref{sec:pulsar_term} in order to estimate $\boldsymbol{\theta}$ and compare the results with those using Earth-terms model, Equation \eqref{eq:g_func_trig_earth}. In Section \ref{sec:detection_new} we calculate the detectability of the GW source as a function of $h_0$ for both the pulsar terms and the Earth terms model. \newline

This section considers only a single realisation of the system (i.e. process and measurement) noise and a particular, arbitrary set of injection parameters $\boldsymbol{\theta}$. Extension to multiple noise realisations and broader parameter spaces is considered in Sections X and Y respectively.

\subsection{Generating synthetic data}\label{sec:synthetic_data}
\begin{table*}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lccll}
			\toprule
			&Parameter & Injected Values & Units & Prior  \\
			\hline
			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
			& $\alpha$          & $1.0$  & radians & Uniform($0, 2 \pi $)\\
			& $\delta$              & $1.0$  & radians & Uniform($-\pi/2, \pi/2$) \\
			& $\psi$              & $2.50$ & radians & Uniform($0, 2 \pi $) \\
			& $\Phi_0$          & $0.20$ & radians & Uniform($0, 2 \pi $) \\
			& $h_0$            & $5 \times 10^{-15}$ & - & LogUniform($10^{-15}$, $10^{-9}$) \\
			& $\iota$             & $1.0$ & radians & Uniform($0, \pi$) \\ 
			\hline
			\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left( f_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{f} \right)$ \\
			& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left( \dot{f}_{\rm ATNF}^{(n)} - 10^3 \epsilon^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \epsilon^{(n)}_{\dot{f}} \right)$ \\
			&  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & - \\
			& $\sigma^{(n)}$              & $\sigma_{\rm sc}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & - \\
			\bottomrule
		\end{tabular}
		\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the choice of prior used for Bayesian inference on each parameter. The subscript ``ATNF" denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}. The subscript ``sc" indicates that the injected value has been calculated using the empirical timing model, Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} from \protect \cite{Shannon2010}. The quantities $\epsilon^{(n)}_{f}$ and $\epsilon^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue.}
		\label{tab:parameters_and_priors}
	\end{table*}

>>>>

The modulation in the pulsar TOAs due to the presence of a GW has two contributions: an ``Earth term" and a ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. Whilst the Earth term is phase coherent between all pulsars, the pulsar terms have uncorrelated phases. Consequently the pulsar terms are typically considered as a source of self-noise and dropped from many standard PTA analyses \citep[e.g.][]{Sesana2010,Babak2012,Petiteau2013,Zhu2015,Taylors2016,Goldstein2018,Charisi2023arXiv230403786C}. However, dropping the pulsar terms leads to biases in the inferred parameters, and reduces the detection probability \citep{Zhupulsarterms,Chen2022,KimpsonPTA}. \newline

\cite{KimpsonPTA} (K24 hereafter) introduced a novel, state-space method for the detection and parameter estimation of continuous GWs from individual SMBHBs, that self-consistently tracks the intrinsic timing noise in PTA pulsars \citep[e.g.][]{Shannon2010,Lasky2015,Caballero2016,Goncharov2021} and disentangles it from GW-induced TOA modulations. The method described in K24 follows the example of standard  PTA analyses and also drops the pulsar term. In this paper we undertake two key tasks. Firstly we extend the methods presented in K24 to include the pulsar term. We show how to reparametrise the model in terms of an additional pulsars-specific phase term. We demonstrate how the static GW parameters and the additional pulsar phase term can be estimated by combining a Kalman filter \citep{Kalman1} with a Bayesian nested sampler \citep{Skilling, Ashton2022}. Secondly we quantify how the inclusion of the pulsar term affects the accuracy with which the static GW parameters can be estimated, and the minimal detectable GW strain. \newline

The paper is organised as follows. In Section \ref{sec:2} we briefly review the state-space model of PTA data analysis introduced in \cite{KimpsonPTA}. In Section \ref{sec:pulsar_term} we show how to include the pulsar term in the model via a convenient reparametrisation. In Section \ref{sec:rep_example1} we test the updated model which includes the pulsar term on synthetic data for a single representative GW source. In Section \ref{sec:earth_vs_psr} we compare the parameter estimation accuracy and the minimal detectable GW strain when including the pulsar terms to the case where the pulsar terms are dropped. In Section \ref{sec:parameter_space} we extend the tests of Section \ref{sec:rep_example1} to cover an astrophysically relevant domain of SMBHB source parameters. Conclusions are drawn in Section \ref{sec:discussion}. Throughout the paper we adopt the natural units, with $c = G = \hbar = 1$, and metric signature $(-,+,+,+)$. \newline

@@

%The paper is organised as follows. In Section \ref{sec:model} we present the state-space model for the rotational states of an array of pulsars falling freely in the curved spacetime of a single-source GW. In Section \ref{sec:detect} we develop a Kalman filter to track the state evolution and deploy the Kalman filter in conjunction with nested sampling to estimate the GW and other system parameters, along with the model evidence. In Section \ref{sec:testing} we describe how we create synthetic validation data with which to test the method. We go on in Section \ref{sec:rep_example} to test the method on the synthetic data for a single representative GW source. In Section \ref{sec:parameter_space} we extend the tests to cover an astrophysically relevant domain of SMBHB source parameters. In Section \ref{sec:bias_and_identifiability} we discuss identifiability of and bias in the parameter estimates.

@@

There are $N$ pulsars in the array. The index $n$ labels the $n$-th pulsar, with $1\leq n\leq N$. The intrinsic spin frequency of the $n$-th pulsar as measured in the local, freely-falling rest frame of the pulsar's centre of mass, $f_{\rm p}^{(n)}(t)$, evolves according to a stochastic differential equation. The radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$, is related to $f_{\rm p}^{(n)}(t)$ via a measurement equation which quantifies the influence of the GW. In Section \ref{sec:psr_frequency} we outline the phenomenological model for the evolution of $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:psr_measured} we outline the measurement equation which relates $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$. In Section \ref{sec:ss_params} we summarise the parameters of the model. The model is formulated in terms of pulse frequency time series as a proof of principle and to maintain consistency with previous work \citep{Myers2021MNRAS.502.3113M,Meyers2021,KimpsonPTA}. It will be necessary to modify the method to accept pulse TOAs instead of a pulse frequency time series when implementing it to analyze real data, a subtle generalization we defer to future work.

\subsection{Spin evolution} \label{sec:psr_frequency}
The intrinsic pulsar spin frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas},
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)}(t) +\xi^{(n)}(t) \ , 
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm em}^{(n)}$ is the deterministic evolution, $\dot{f}_{\rm em}^{(n)}$ is the spin derivative, $\gamma^{(n)}$ is a damping constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ is a white noise stochastic process which satisfies:
\begin{align}
	\langle \xi^{(n)}(t) \rangle &= 0 \ , \\
	\langle \xi^{(n)}(t) \xi^{(n')}(t') \rangle &= [\sigma^{(n)}]^2 \delta_{n,n'}(t - t') \ .	\label{eq:xieqn}
\end{align}
In Equation \eqref{eq:xieqn} $[\sigma^{(n)}]^2$ is the variance of $\xi^{(n)}$ and parametrizes noise amplitude and the gives characteristic root mean square fluctuations $\approx \sigma^{(n)} / [\gamma^{(n)}]^{1/2}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. The deterministic evolution $f_{\rm em}^{(n)}$ is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$ \citep{1969ApJ...157..869G}. PTAs are typically composed of millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and can be approximated accurately by 
\begin{equation}
	f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} 
where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. \newline

@@

\section{Inclusion of the pulsar term}\label{sec:pulsar_term}
K24 uses a likelihood-based Bayesian framework for inference of the model parameters and model selection. That is, given a temporal sequence of noisy measurements $\boldsymbol{Y}(t)$, the posterior distribution of $\boldsymbol{\theta}$ is calculated by Bayes' Rule,
\begin{equation}
	p(\boldsymbol{\theta} | \boldsymbol{Y}) = \frac{\mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta}) \cdot \pi(\boldsymbol{\theta})}{\mathcal{Z}} \ ,
\end{equation}
where $\pi(\boldsymbol{\theta})$ is the prior distribution on $\boldsymbol{\theta}$ and $\mathcal{Z}$ is the marginalised likelihood, or evidence
\begin{equation}
	\mathcal{Z} = \int d \boldsymbol{\theta} \mathcal{L}(\boldsymbol{Y} | \boldsymbol{\theta})  \pi(\boldsymbol{\theta})  \ . \label{eq:model_evidence}
\end{equation}
and $\mathcal{L}(\boldsymbol{Y}| \boldsymbol{\theta})$ is the likelihood calculated via a Kalman filter \citep{Kalman1}. For our purposes ${\boldsymbol{Y}}(t) = \{ f_{\rm m}^{(n)}(t) \}$. Nested sampling \citep{Skilling} is used to estimate $p(\boldsymbol{\theta} | \boldsymbol{Y})$ and $\mathcal{Z}$. We refer the reader to Appendix X for a review of Kalman filtering, nested sampling and the combined workflow in the context of PTA searches for GWs. \newline

@@

\subsection{Reparametrisation of the pulsar term} \label{sec:pulsar_term2}
We define a new parameter 
\begin{equation}
	\chi^{(n)} = \Omega \left( 1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \mod 2 \pi 
\end{equation}
Such that Equation \eqref{eq:g_func_trig} becomes
\begin{align}
	g^{(n)}(t) =& 1 - \frac{ H_{ij}[q^{(n)}]^i [q^{(n)}]^j }{2 [1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}] } \nonumber \\
	& \times \Big[\cos\left(-\Omega t +\Phi_0\right) \nonumber \\
	&- \cos \left \{-\Omega t +\Phi_0 + \chi^{(n)} \right \} \Big ] \ .
	\label{eq:g_func_trig_chi}
\end{align}
and the static pulsar parameters to be inferred become
\begin{equation}
	\boldsymbol{\theta'}_{\rm psr} = \left \{ \gamma^{(n)},\sigma^{(n)}, f_{\rm em}^{(n)}(t_1),\dot{f}_{\rm em}^{(n)}(t_1),\chi^{(n)}\right\}_{1\leq n \leq N} \ .  \label{eq:psrparams}
\end{equation}
whilst $\boldsymbol{\theta}_{\rm gw}$  remain unchanged. We define $\boldsymbol{\theta'} =  \boldsymbol{\theta}_{\rm psr} \cup \boldsymbol{\theta'}_{\rm gw}$. \newline

@@

\section{Representative example: single SMBHB source including the pulsar terms}\label{sec:rep_example1}
In this section we apply the analysis scheme of K24 while including the pulsar terms using the parametrisation described in Section \ref{sec:pulsar_term2}. As outlined in Section \ref{sec:pulsar_term}, the analysis scheme consists of a Kalman filter to track the state evolution of the intrinsic pulsar spin frequency which provides a likelihood for nested sampling. Nested sampling then returns estimates of the static parameters of the model and $\mathcal{Z}$. We refer the reader to Appendix X and K24 for details of the Kalman filter - nested sampling scheme. We apply the analysis scheme to a PTA perturbed by a GW from an individual quasi-monochromatic SMBHB source. The analysis of a stochastic background composed of the superposition of multiple sources is more challenging and is postponed to a forthcoming paper. The goal of this section is to validate the analysis scheme when the pulsar terms are included. \newline

In Section \ref{sec:synthetic_data} we demonstrate how to generate synthetic data, namely noisy frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$, with which to validate the scheme. In Section \ref{sec:pe_and_ms} we apply the analysis scheme to the synthetic data to estimate $\boldsymbol{\theta'}$ and $\mathcal{Z}$. The scheme is validated on multiple noise realisations of the pulsar process noise and the measurement noise in order to test the scheme more strongly and quantify the  irreducible ``cosmic'' variance in the parameter estimates. This section considers only a particular, arbitrary set of injection parameters $\boldsymbol{\theta'}$. Extension to broader parameter spaces is considered in Section X.

\subsection{Generating synthetic data}\label{sec:synthetic_data}
\begin{table*}
	\centering
	%	\resizebox{\textwidth}{!}{%
		%	\renewcommand{\arraystretch}{1.0} % Default value: 1
		\begin{tabular}{lccll}
			\toprule
			&Parameter & Injected value & Units & Prior  \\
			\hline
			\multirow{7}{2mm}{$\boldsymbol{\theta}_{\rm gw}$} & $\Omega$       & $5 \times 10^{-7}$ & Hz & LogUniform($10^{-9}$, $10^{-5}$) \\
			& $\alpha$          & $1.0$  & rad & Uniform($0, 2 \pi $)\\
			& $\delta$              & $1.0$  & rad & Uniform($-\pi/2, \pi/2$) \\
			& $\psi$              & $2.50$ & rad & Uniform($0, 2 \pi $) \\
			& $\Phi_0$          & $0.20$ & rad & Uniform($0, 2 \pi $) \\
			& $h_0$            & $5 \times 10^{-15}$ & --- & LogUniform($10^{-15}$, $10^{-9}$) \\
			& $\iota$             & $1.0$ & rad & Uniform($0, \pi$) \\ 
			\hline
			%		\vspace{1mm}\multirow{5}{2mm}{$\boldsymbol{\theta}_{\rm psr}$} & $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
			& $f_{\rm em}^{(n)} (t_1)$       & $f_{\rm ATNF}^{(n)}$ & Hz & Uniform$\left[f_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{f}, f_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{f} \right]$ \\
			& $\dot{f}_{\rm em}^{(n)} (t_1)$       & $\dot{f}_{\rm ATNF}^{(n)}$ & s$^{-2}$ & Uniform$\left[ \dot{f}_{\rm ATNF}^{(n)} - 10^3 \eta^{(n)}_{\dot{f}}, \dot{f}_{\rm ATNF}^{(n)} + 10^3 \eta^{(n)}_{\dot{f}} \right]$ \\
			\vspace{1mm} $\boldsymbol{\theta}_{\rm psr}$ &  $d^{(n)}$       &$d_{\rm ATNF}^{(n)}$  & m & --- \\
			\vspace{1mm} & $\sigma^{(n)}$              & $\sigma_{\rm SC}^{(n)}$ & $s^{-3/2}$ & LogUniform($10^{-23}, 10^{-19}$) \\
			& $\gamma^{(n)}$              & $10^{-13}$ & s$^{-1}$ & --- \\
			\bottomrule
		\end{tabular}
		\caption{Summary of injected static parameters used for generating synthetic data in the representative example of Section \ref{sec:rep_example}, along with the prior used for Bayesian inference on each parameter (rightmost column). The top and bottom halves of the table contain the values for $\boldsymbol{\theta}_{\rm gw}$ and $\boldsymbol{\theta}_{\rm psr}$ respectively.The subscript ``ATNF'' denotes values which have been obtained from the ATNF pulsar catalogue as described in Section \ref{sec:synt_pta}. The subscript ``SC'' indicates that the injected value has been calculated using Equations \eqref{eq:sigmap} and \eqref{eq:sigmap_f} in \protect \cite{Shannon2010}. The quantities $\eta^{(n)}_{f}$ and $\eta^{(n)}_{\dot{f}}$ are the errors in $f^{(n)}_{\rm em} (t_1)$ and $\dot{f}^{(n)}_{\rm em} (t_1)$ respectively, as quoted in the ATNF catalogue. No priors are set for $d^{(n)}$ and $\gamma^{(n)}$ since those parameters do not enter the inference model,  as discussed in Section \ref{sec:priors}, c.f. Equation \eqref{eq:psrreduced}.
		}
		\label{tab:parameters_and_priors}
	\end{table*}
In order to generate synthetic frequency time series $f_{\rm m}^{(1)}(t), \dots, f_{\rm m}^{(N)}(t)$ Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} are integrated numerically using a Runge-Kutta It$\hat{\text{o}}$ integrator implemented in the \texttt{sdeint} python package \footnote{\url{https://github.com/mattja/sdeint}}. This produces random realizations of $f_{\rm p}^{(n)}(t)$ for $1\leq n \leq N$. The measured $f_{\rm m}^{(n)}(t)$ are then produced via Equations \eqref{eq:measurement}--\eqref{eq:g_func_trig}. Equations \eqref{eq:frequency_evolution}--\eqref{eq:spinevol} and \eqref{eq:measurement}--\eqref{eq:g_func_trig} depend on the choice of static parameters $\boldsymbol{\theta'}$, as well as the choice of pulsar directions ${\boldsymbol{q}}^{(n)}$ and detector measurement noise covariance $\sigma_{\rm m}$ (Equation \eqref{eq:sigma_m_eqn}). We now describe how these values are chosen so as to generate astrophysically representative synthetic data. \newline

In this paper we use the same values for $\boldsymbol{\theta}_{\rm psr}$ as in K24. That is, we adopt the 47 millisecond pulsars (MSPs) in the 12.5 year NANOGrav dataset \citep{2020ApJ...905L..34A}, i.e. $N = 47$. We consider all pulsars to be observed with a weekly cadence, $T_{\rm cad} = 1 \,{\rm week}$ over a 10 year period. For each pulsar fiducial values for ${\boldsymbol{q}}^{(n)}$, $d^{(n)}$, $f_{\rm em}^{(n)}(t_1)$, and $\dot{f}^{(n)}_{\rm em}(t_1)$ are obtained via the Australia Telescope National Facility (ATNF) pulsar catalogue \citep{Manchester2005} using the \texttt{psrqpy} package \citep{psrqpy}. We identify $f_{\rm em}^{(n)} (t_1)$ and $\dot{f}_{\rm em}^{(n)} (t_1)$ with the pulsar barycentric rotation frequency and its time derivative respectively, as quoted in catalogues. No direct measurements exist for $\gamma^{(n)}$, though the mean reversion timescale typically satisfies $[\gamma^{(n)}]^{-1} \gg T_{\rm obs}$ \citep{Price2012,Myers2021MNRAS.502.3113M,Meyers2021,Vargas}. In this paper, for the sake of simplicity in the absence of independent measurements, we fix $\gamma^{(n)} = 10^{-13}$ s$^{-1}$ for all $n$. Similarly, no direct measurements exist for $\sigma^{(n)}$.   We relate the process noise in the frequency $\sigma^{(n)}$ to the root mean square TOA noise accumulated over one week such that
\begin{eqnarray}
	\sigma^{(n)} \approx \sigma_{\rm TOA}^{(n)} f_{\rm p}^{(n)} {T_{\rm cad}}^{-3/2} \ . \label{eq:sigmap_f}
\end{eqnarray}
where, following K24, we use the empirical timing noise model for MSPs from \cite{Shannon2010ApJ...725.1607S} in order to set physically reasonable values on $\sigma_{\rm TOA}^{(n)}$.  The median $\sigma^{(n)}$ calculated in this way is $\sigma^{(n)} = 5.51 \times 10^{-24} $ s$^{-3/2}$, $\min [ \sigma^{(n)} ] = 1.67 \times 10^{-26}$s$^{-3/2}$ for PSR J0645+5158 and $\max [ \sigma^{(n)} ] = 2.56 \times 10^{-19}$ s$^{-3/2}$ for PSR J1939+2134. Compared to the other 46 pulsars of the array, PSR J1939+2134 is a notable outlier with an estimated $\sigma^{(n)}$ much larger than the median. In practice, in order to avoid setting an overly broad prior on $\sigma^{(n)}$ (see Section \ref{sec:set_priors}) for validation purposes, we artificially set $\sigma^{(n)} = 10^{-20} \, {\rm s^{-3/2}}$ for this pulsar, so that $\sigma^{(n)}$ for every pulsar in the synthetic PTA falls within an reasonable width log-uniform prior. This is done purely for testing convenience; it is straightforward to expand the prior when analysing real, astronomical data. \newline
